### 2016-07-10 Projects/DelCo_2012/Report/DelCo_2012.tex
minor report edits, add data file and script for making scatters

\begingroup
{\bf This publication is available free of charge from:}
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.1838}{\bf http://dx.doi.org/10.6028/NIST.TN.1838} \\
\endgroup

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel member. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood based materials.  The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished.  Wood based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan.  In order to accomplish the last series of experiments, additional funding was obtained from NIST and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets as well as carpeting and upholstered furniture.

@@

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior.  The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure.  When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure.  This was done to minimize, the build-up of "dry" foam or a slug upon the initial flow of the nozzle into the test structure.

@@

To better quantify the effectiveness of the suppression agents (CAF, water), the maximum temperature differences associated with each gas cooling event for each agent from the three thermocouple trees were analyzed. Figure~\ref{fig:combined_all} compares the temperature differences at each thermocouple position (via a scatterplot) for 30 of the 88 experiments.  The 30 experiments are actually pairs of experiments, 15 water and 15 CAF, which were directly comparable, see Table~\ref{tab:Gas_Cooling_Descriptions}.  These experiments included solid stream, straight streams, and narrow fog streams at both nozzle positions.  While the total number of experiments was 88, there where a number of experiments that were performed which did not have a directly comparable experiments between water and CAF.  For example a set of 12 Class A solutions experiments were conducted.  Gas cooling experiments were conducted with water applied on a wide fog nozzle setting, no significant differences were noted so a comparable series with wide fog CAF was not conducted.  Similarly larger water flows were examined but 120 gpm was the maximum water flow that could be generated with the CAFS, so those experiments are not included in this comparison.  The Mitchell report does include all of the data for the 88 experiments \cite{Mitchell:1}

@@

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Replicate Tests}
	\label{fig:combined_all}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_mid_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Middle Position Replicates}
	\label{fig:CAFS_Water_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Back Position Replicates}
	\label{fig:CAFS_Water_full}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/PDFs/West_Structure/DelCo_2012_West_Structure_Instrumentation2}
	\caption{Schematic of  the Gas Cooling Experiments.}
	\label{fig:Gas_Cooling_Instrumentation_Dimensions2}
\end{figure}
Based on the monitor position (arrow at the doorway in Fig.~\ref{fig:Gas_Cooling_Instrumentation_Dimensions2}), thermocouple arrays A1 and A2 had more direct impact from the suppression agent compared to thermocouple array A3. As an example, Figure~\ref{fig:Burn_Building_Test_18b} shows the distribution of the agent within the burn building from a spray density test for a straight stream with CAF at the mid position compared to the positions of the thermocouple trees.
\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Bars/BB18b}
	\caption{Gas Cooling Spray Density Test 18, CAF, Straight Stream}
	\label{fig:Burn_Building_Test_18b}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A1 Experimental Pairs}
	\label{fig:CAFS_Water_A1_all}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A2 Experimental Pairs}
	\label{fig:CAFS_Water_A2_all}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A3 Experimental Pairs}
	\label{fig:CAFS_Water_A3_all}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A1 Replicates}
	\label{fig:CAFS_Water_A1_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A2 Replicates}
	\label{fig:CAFS_Water_A2_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A3 Replicates}
	\label{fig:CAFS_Water_A3_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A1 Replicates}
	\label{fig:CAFS_Water_A1_back}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A2 Replicates}
	\label{fig:CAFS_Water_A2_back}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A3 Replicates}
	\label{fig:CAFS_Water_A3_back}
\end{figure}

@@

There were 12 suppression tests conducted in the single story purpose-built concrete structure. Eight of the experiments used a fuel load comprised of typical household furnishings that included wooden wall paneling, two sofas, two overstuffed chairs, carpet, and padding (Section~\ref{sec:fire_suppression_furniture_fuel}). The other four experiments used wood pallet fuel load in a space with the ceiling and walls composed of wood based materials (Section~\ref{sec:fire_suppression_pallet_fuel}). Ignition occurred in the rear of the single story structures and the available ventilation was through the front doorway and for eight of the experiments a rear window was "opened" to provide additional ventilation (see Table~\ref{tab:Test_Descriptions}). The fires were intended to replicate post-flashover residential room and contents fires. Hose lines were pre-deployed with fixed nozzles set in two locations: one in the hallway on the approach to the burn room (hallway nozzle) and one just inside the burn room to the right side of the hallway (room nozzle). Each nozzle flowed 15~s of water or compressed air foam depending on the test. The hallway nozzle was intended to simulate a crew advancing into a structure towards the burn room. Suppression from this location would be in the form of gas cooling as the stream was not aimed at the seat of the fire (indirect interior attack).  After the hallway nozzle was shutdown, there was a 15~s time period to simulate the crew advancing into the doorway of the fire room. The room nozzle was positioned just to the side of the entry door.  It was also opened and flowed for approximately 15~s. The room nozzles was aimed in the direction of the seat of the fire (direct interior attack). Table~\ref{tab:Test_Descriptions} provides an overview of the suppression tests conducted.

@@

\subsection{Iowa State Univeristy Fire Flow Formula (1963)~\cite{NFPA}}
The Iowa Formula for determining the required fire flowrate for a given structure was developed, by Keith Royer and Bill Nelson, based on theoretical heat absorption capabilities of water. The required fire flowrate in gpm is equal to the volume of the fire compartment (in cubic feet) divided by 100:
\begin{equation}\label{eq:isu_form}
required fire flowrate = Volume / 100
\end{equation}
Due to inefficiencies in applying the water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the  based on floor area yields an application density of 3.3~Lpm/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~Lpm/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). Our fire rooms ranged in floor area from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$).  Therefore our required flowrate for the fire room in the single story structure based on the Iowa State University formula would be approximately 87~Lpm (23~gpm).  The required flowrate for the fire room in the lower level of the two story would be approximately 168~Lpm (44~gpm).

\subsection{National Fire Academy (1884)~\cite{Klaene:1}}
The National Fire Academy (NFA) developed a formula based on a study of fire flows, which were successful in controlling a large number of working fires~\cite{Klaene:1}. This formula sets the required fire flow, in gpm, equal to the floor area (in square feet) divided by 3.
\begin{equation}\label{eq:nfa_form}
required fire flowrate = Area / 3
\end{equation}

Our fire rooms ranged in floor area from approximately 27~$m^2$ (291~$ft^2$) to 51.4~$m^2$ (553~$ft^2$).
Applying the NFA formula based on floor area yields an required flowrate for the single story fire room of 369~Lpm (97~gpm).  The required flowrate for the fire room in the lower level of the two story would be approximately 700~Lpm (184~gpm).

>>>>

\begingroup
{\bf This publication is available free of charge from:}
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.XXXX}{\bf http://dx.doi.org/10.6028/NIST.TN.XXXX} \\
\endgroup

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel member. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood based materials. The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. Wood based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets as well as carpeting and upholstered furniture.

@@

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure. This was done to minimize, the build-up of``dry'' foam or a slug upon the initial flow of the nozzle into the test structure.

@@

To better quantify the effectiveness of the suppression agents (CAF, water), the maximum temperature differences associated with each gas cooling event for each agent from the three thermocouple trees were analyzed. Figure~\ref{fig:combined_all} compares the temperature differences at each thermocouple position (via a scatterplot) for 30 of the 88 experiments. The 30 experiments are actually pairs of experiments, 15 water and 15 CAF, which were directly comparable, see Table~\ref{tab:Gas_Cooling_Descriptions}.  These experiments included solid stream, straight streams, and narrow fog streams at both nozzle positions.  While the total number of experiments was 88, there where a number of experiments that were performed which did not have a directly comparable experiments between water and CAF. For example a set of 12 Class A solutions experiments were conducted. Gas cooling experiments were conducted with water applied on a wide fog nozzle setting, no significant differences were noted so a comparable series with wide fog CAF was not conducted. Similarly larger water flows were examined but 120 gpm was the maximum water flow that could be generated with the CAFS, so those experiments are not included in this comparison. The Mitchell report includes the data for all 88 experiments \cite{Mitchell:1}

@@

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Replicate Tests}
	\label{fig:combined_all}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Middle Position Replicates}
	\label{fig:CAFS_Water_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Back Position Replicates}
	\label{fig:CAFS_Water_full}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/PDFs/West_Structure/DelCo_2012_West_Structure_Instrumentation2}
	\caption{Schematic of  the Gas Cooling Experiments.}
	\label{fig:Gas_Cooling_Instrumentation_Dimensions2}
\end{figure}
Based on the monitor position (arrow at the doorway in Fig.~\ref{fig:Gas_Cooling_Instrumentation_Dimensions2}), thermocouple arrays A1 and A2 had more direct impact from the suppression agent compared to thermocouple array A3. As an example, Figure~\ref{fig:Burn_Building_Test_18b} shows the distribution of the agent within the burn building from a spray density test for a straight stream with CAF at the mid position compared to the positions of the thermocouple trees.
\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Bars/BB18b}
	\caption{Gas Cooling Spray Density Test 18, CAF, Straight Stream}
	\label{fig:Burn_Building_Test_18b}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A1 Experimental Pairs}
	\label{fig:CAFS_Water_A1_all}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A2 Experimental Pairs}
	\label{fig:CAFS_Water_A2_all}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for All Thermocouple Tree A3 Experimental Pairs}
	\label{fig:CAFS_Water_A3_all}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A1 Replicates}
	\label{fig:CAFS_Water_A1_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A2 Replicates}
	\label{fig:CAFS_Water_A2_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_mid_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Middle Position Thermocouple Tree A3 Replicates}
	\label{fig:CAFS_Water_A3_mid}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A1_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A1 Replicates}
	\label{fig:CAFS_Water_A1_back}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A2_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A2 Replicates}
	\label{fig:CAFS_Water_A2_back}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.7\columnwidth]{../Figures/Gas_Cooling/Combined_fullback_A3_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for Back Position Thermocouple Tree A3 Replicates}
	\label{fig:CAFS_Water_A3_back}
\end{figure}

@@

There were 12 suppression tests conducted in the single story purpose-built concrete structure. Eight of the experiments used a fuel load comprised of typical household furnishings that included wooden wall paneling, two sofas, two overstuffed chairs, carpet, and padding (Section~\ref{sec:fire_suppression_furniture_fuel}). The other four experiments used wood pallet fuel load in a space with the ceiling and walls composed of wood based materials (Section~\ref{sec:fire_suppression_pallet_fuel}). Ignition occurred in the rear of the single story structures and the available ventilation was through the front doorway and for eight of the experiments a rear window was ``opened''to provide additional ventilation (see Table~\ref{tab:Test_Descriptions}). The fires were intended to replicate post-flashover residential room and contents fires. Hose lines were pre-deployed with fixed nozzles set in two locations: one in the hallway on the approach to the burn room (hallway nozzle) and one just inside the burn room to the right side of the hallway (room nozzle). Each nozzle flowed 15~s of water or compressed air foam depending on the test. The hallway nozzle was intended to simulate a crew advancing into a structure towards the burn room. Suppression from this location would be in the form of gas cooling as the stream was not aimed at the seat of the fire (indirect interior attack).  After the hallway nozzle was shutdown, there was a 15~s time period to simulate the crew advancing into the doorway of the fire room. The room nozzle was positioned just to the side of the entry door.  It was also opened and flowed for approximately 15~s. The room nozzles was aimed in the direction of the seat of the fire (direct interior attack). Table~\ref{tab:Test_Descriptions} provides an overview of the suppression tests conducted.

@@

\subsection{Iowa State Univeristy Fire Flow Formula (1963)}
The Iowa Formula for determining the required fire flowrate for a given structure was developed, by Keith Royer and Bill Nelson, based on theoretical heat absorption capabilities of water. The required fire flowrate in gpm is equal to the volume of the fire compartment (in cubic feet) divided by 100:
\begin{equation}\label{eq:isu_form}
required fire flowrate = Volume / 100
\end{equation}
Due to inefficiencies in applying the water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the  based on floor area yields an application density of 3.3~Lpm/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~Lpm/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). Our fire rooms ranged in floor area from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). Therefore the required flowrate for the fire room in the single story structure based on the Iowa State University formula would be approximately 87~Lpm (23~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 168~Lpm (44~gpm).

\subsection{National Fire Academy (1884)}
The National Fire Academy (NFA) developed a formula based on a study of fire flows, which were successful in controlling a large number of working fires~\cite{Klaene:1}. This formula sets the required fire flow, in gpm, equal to the floor area (in square feet) divided by 3.
\begin{equation}\label{eq:nfa_form}
required fire flowrate = Area / 3
\end{equation}

The fire rooms in these experiments ranged in floor area from approximately 27~$m^2$ (291~$ft^2$) to 51.4~$m^2$ (553~$ft^2$). Applying the NFA formula based on floor area yields an required flowrate for the single story fire room of 369~Lpm (97~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 700~Lpm (184~gpm).
